---
title: "Data Analyst position at Raków Częstochowa - Task"
author: "Dr. Artur Czeszumski"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cerulean
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, results = 'hide')
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggsoccer)
  # Add other libraries as needed
})
```

W tym dokumencie przedstawiam odpowiedzi na pytania zwiazane z aplikacja na pozycje analityka danych w zespole Raków Częstochowa. Ten dokument zawiera kod, wizualizacje i odpowiedzi na pytania. Komentarze do kodu są w języku angielskim.

```{r, warning=FALSE}
# Load required libraries
library(jsonlite)
library(tidyverse)
library(knitr)
library(ggsoccer)

# Read the JSON files. Please remember to keep this file in the same folder as the data shared for the assignment. 
d_match_stats <- stream_in(file('player_match_stats_match_3891414.json'))
d_events <- stream_in(file('events_match_3891414.json'))
d_lineup <- stream_in(file('lineups_match_3891414.json'))
```

# Pytanie 1:

Oblicz różnicę xG (expected goals) dla każdego stanu meczu (wygrana-przegrana-remis) dla obu drużyn.

```{r, results = 'asis'}
# Filter events with xG values (shots)
shots_xg <- !is.na(d_events$shot$statsbomb_xg)

# Create a dataframe with xG values, outcomes, teams, and minutes of each shot
xG <- data.frame(d_events[shots_xg,]$shot$statsbomb_xg, 
                 d_events[shots_xg,]$shot$outcome$name,
                 d_events[shots_xg,]$possession_team$name,
                 d_events[shots_xg,]$minute)
# change column names for more readable
colnames(xG) <- c('xG', 'Outcome', 'Team', 'Minute') 
# create a variable that keeps both team names
teams <- unique(xG$Team)
# Add initial row with minute 0 and accumulated xG of 0 for each team
initial_rows <- data.frame(
  xG = 0,
  Outcome = NA,
  Team = teams,
  Minute = 0
)
# combine empty row with the data frame
xG <- rbind(initial_rows, xG)

# Convert Team to a factor and reverse the levels
xG$Team <- factor(xG$Team)
xG$Team <- factor(xG$Team, levels = rev(levels(xG$Team)))

# Calculate accumulated xG for each team over time
xG <- xG %>%
  arrange(Team, Minute) %>%
  group_by(Team) %>%
  mutate(accumulated_xG = cumsum(xG))

# Find the maximum minute in the match
max_minute <- max(xG$Minute)

# Add final row with maximum minute for each team
final_rows <- xG %>%
  group_by(Team) %>%
  summarize(Minute = max_minute, accumulated_xG = last(accumulated_xG), .groups = 'drop') %>%
  mutate(xG = 0, Outcome = NA)
xG <- rbind(xG, final_rows)

# Filter data for goals
goals <- xG %>%
  filter(Outcome == "Goal")

# Calculate the score for each team
score <- goals %>%
  group_by(Team) %>%
  summarise(Goals = n()) %>%
  ungroup()

# Ensure both teams are in the score data frame
score <- complete(score, Team, fill = list(Goals = 0))

# Create a label for the game score
game_score_label <- paste(score$Team[1], score$Goals[1], "-", score$Goals[2], score$Team[2])

# Plot accumulated xG over time for each team using step lines
ggplot(xG, aes(x = Minute, y = accumulated_xG, color = Team, group = Team)) +
  geom_step(size = 1.5) +  # Draw step lines to show the progression of accumulated xG over time
  geom_point(data = goals, aes(x = Minute, y = accumulated_xG, fill = Team), size = 3, shape = 21) +  # Add points for goals, using a filled circle
  geom_text(data = goals, aes(x = Minute, y = accumulated_xG, 
                              label = paste0("xG: ", sprintf("%.2f", xG))), 
            vjust = -1, hjust = 0.85, size = 4, color = "black") +  # Annotate goals with their xG values, positioned above the points
  geom_text(data = final_rows, aes(x = Minute, y = accumulated_xG, 
                                   label = sprintf("%.2f", accumulated_xG), color = Team), 
            hjust = -0.3, size = 6, show.legend = FALSE) +  # Annotate the final accumulated xG values for each team, positioned slightly to the right of the last point
  labs(title = game_score_label,  # Set the title of the plot
       x = "Minuta",  # Label for the x-axis
       y = "xG") +  # Label for the y-axis
  scale_y_continuous(breaks = seq(0, max(xG$accumulated_xG) * 1.2, by = 0.25)) +  # Set y-axis breaks every 0.25 units, extending 20% beyond the max value
  scale_x_continuous(breaks = seq(0, max(xG$Minute), by = 15)) +  # Set x-axis breaks every 15 minutes
  scale_color_brewer(palette = "Set1") +  # Apply a color palette from ColorBrewer to the lines and points
  scale_fill_brewer(palette = "Set1") +   # Ensure the fill color for points matches the line color
  coord_cartesian(clip = 'off') +  # Ensure text annotations are not clipped by the plot area
  theme_minimal() +  # Use a minimal theme for a clean look
  theme(legend.title = element_blank(),  # Remove the legend title
        legend.position = "right",  # Position the legend on the right side of the plot
        legend.text = element_text(size = 14),  # Increase the legend text size
        legend.key.size = unit(1.5, "lines"),   # Increase the size of the legend keys
        title = element_text(size = 14),  # Set the title text size
        plot.margin = unit(c(1, 2, 1, 1), "lines"))  # Adjust the plot margins

### Calculate the difference between xG for each score change


# Function to calculate xG difference at the time of each goal
calculate_xG_difference <- function(goals, xG) {
  results <- data.frame()
  
  for (i in 1:nrow(goals)) {
    goal_event <- goals[i, ]
    goal_minute <- goal_event$Minute
    scoring_team <- goal_event$Team
    
    # Get the state of the game at the goal minute for both teams
    game_state <- xG %>%
      filter(Minute <= goal_minute) %>%
      group_by(Team) %>%
      summarise(accumulated_xG = last(accumulated_xG)) %>%
      ungroup()
    
    # Count the goals up to the goal minute for both teams
    goal_state <- goals %>%
      filter(Minute <= goal_minute) %>%
      group_by(Team) %>%
      summarise(goals_scored = n()) %>%
      ungroup()
    
    # Ensure both teams are included in the game state
    if (nrow(game_state) == 2) {
      team1 <- game_state$Team[1]
      team2 <- game_state$Team[2]
      xg_team1 <- game_state$accumulated_xG[1]
      xg_team2 <- game_state$accumulated_xG[2]
      
      goals_team1 <- ifelse(team1 %in% goal_state$Team, goal_state$goals_scored[goal_state$Team == team1], 0)
      goals_team2 <- ifelse(team2 %in% goal_state$Team, goal_state$goals_scored[goal_state$Team == team2], 0)
      
      difference <- abs(xg_team1 - xg_team2)
      
      # Determine the state of the game for each team based on goals
      if (goals_team1 == goals_team2) {
        state_team1 <- "tie"
        state_team2 <- "tie"
      } else {
        state_team1 <- if (goals_team1 > goals_team2) "win" else "lose"
        state_team2 <- if (goals_team2 > goals_team1) "win" else "lose"
      }
      
      result_row <- data.frame(
        Minute = goal_minute,
        Team1 = team1,
        Team2 = team2,
        accumulated_xG_Team1 = xg_team1,
        accumulated_xG_Team2 = xg_team2,
        goals_Team1 = goals_team1,
        goals_Team2 = goals_team2,
        xG_difference = difference,
        state_Team1 = state_team1,
        state_Team2 = state_team2
      )
      
      results <- rbind(results, result_row)
    }
  }
  
  # Order the results by minute
  results <- results %>%
    arrange(Minute)
  
  return(results)
}

# Calculate the xG differences for goals
goal_differences <- calculate_xG_difference(goals, xG)

# remove unnessary columns
goal_differences <- goal_differences %>%
  select(Team1, Team2, xG_difference, state_Team1, state_Team2)

# Print the results
kable(goal_differences)



```

## Odpowiedź:

W odpowiedzi na to pytanie przygotowałem wizualizację pokazującą zmiany xG dla obu drużyn po każdym oddanym strzale. Wszystkie gole są oznaczone z opisem wartości xG, a zaakumulowany xG dla obu drużyn jest przedstawiony. Dodatkowo policzyłem różnicę między zdobytym i straconym xG dla każdego stanu meczu (wygrana-przegrana-remis) i przedstawiłem wszystkie wartości w tabeli.

Raków przegrana: -0.83\
Raków remis: -0.48\
Raków wygrana: 0.33\
Raków wygrana: 0.5

Piast wygrana: 0.83\
Piast remis: 0.48\
Piast przegrana: -0.33\
Piast przegrana: -0.5

# Pytanie 2:

Policz liczbę podań z półprzestrzeni do pola karnego dla obu drużyn.

```{r}
# define half spaces and penalty area
penalty_x <- 102
penalty_y_min <- 18
penalty_y_max <- 62

# divide y (width) by 5 spaces (left, left half-space, middle, right half-space, right)
left_half_spaces_y_min <- 16
left_half_spaces_y_max <- 32
right_half_spaces_y_min <- 48
right_half_spaces_y_max <- 64

# select data
passes <- data.frame(d_events[d_events$type$id == 30,]$possession_team$name,
                     map_dbl(d_events[d_events$type$id == 30,]$location, 1),
                     map_dbl(d_events[d_events$type$id == 30,]$location, 2),
                     map_dbl(d_events[d_events$type$id == 30,]$pass$end_location, 1),
                     map_dbl(d_events[d_events$type$id == 30,]$pass$end_location, 2),
                     d_events[d_events$type$id == 30,]$pass$outcome$name)
colnames(passes) <- c('Team', 'pass_start_x', 'pass_start_y', 'pass_end_x', 'pass_end_y', 'pass_outcome')

# get passes from half spaces to the penalty box for each team
passes_half_spaces <- passes %>%
  filter(pass_start_x < penalty_x,
    (pass_start_y > left_half_spaces_y_min & pass_start_y < left_half_spaces_y_max) | 
           (pass_start_y > right_half_spaces_y_min & pass_start_y < right_half_spaces_y_max),
         pass_end_x > penalty_x & pass_end_x < 120,
         pass_end_y > penalty_y_min & pass_end_y < penalty_y_max)

teams <- unique(passes$Team)

nrow(passes_half_spaces[passes_half_spaces$Team == teams[1],])
nrow(passes_half_spaces[passes_half_spaces$Team == teams[2],])

# Convert Team to a factor and reverse the levels
passes_half_spaces$Team <- factor(passes_half_spaces$Team)
passes_half_spaces$Team <- factor(passes_half_spaces$Team, levels = rev(levels(passes_half_spaces$Team)))

# make a pitch to plot passes
ggplot(passes_half_spaces) +
  annotate_pitch(dimensions = pitch_statsbomb) +
  geom_segment(aes(x = pass_start_x, y = pass_start_y, xend = pass_end_x, yend = pass_end_y, colour = Team),
               arrow = arrow(length = unit(0.25, "cm"),
                             type = "closed")) +
  scale_color_brewer(palette = "Set1") +
  theme_pitch() +
  labs(title = 'Podania z półprzestrzeni do pola karnego')

```

## Odpowiedź:

By zdefiniować półprzestrzenie podzieliłem szerokość boiska na pięć części i wybrałem dwie pomiedzy prawą i środkową, lewa i środkową. Co więcej półprzestrzenia konczą sie na wysokości lini pola karnego.
